Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})

### 2014-05-20 manuscript.tex
Edits to abstract/introduction

The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of lateral organ growth, is a major QTL involved in branching differences between maize and its wild progenitor, teosinte. Previous work has shown that the insertion of a transposable element (\emph{Hopscotch}) upstream of \emph{tb1} enhances the gene's expression, causing much of the reduction in tillering observed in domesticated maize. Observations of the maize \emph{tb1} allele in teosinte individuals, coupled with estimates of an age of insertion of the \emph{Hopscotch} element that pre-dates domestication, led us to investigate its prevalence and potential role in teosinte. Results from genotyping across many natural populations suggest that the \emph{Hopscotch} element is segregating at a higher than expected frequency in a number of populations of two subspecies of teosinte, \emph{Zea mays} ssp. \emph{parviglumis} and \emph{Zea mays} ssp. \emph{mexicana}. Analysis of linkage disequilibrium between the \emph{Hopscotch} element and variation in surrounding regions does not support a hypothesis of recent introgression from maize into teosinte.  Population genetic signatures, however, are consistent with selection on this locus and suggest the \emph{Hopscotch} insertion at \emph{tb1} may play an ecological role in teosinte. Finally, two greenhouse experiments with \emph{Zea mays} ssp. \emph{parviglumis} do not suggest \emph{tb1} controls tillering in natural populations of this subspecies. Our findings suggest that the role of the \emph{Hopscotch} in tillering in teosinte differs from domesticated maize, and that other loci may play a role in observed variation in this trait.

@@

Many morphological changes are associated with maize domestication, and understanding the genetic basis of these changes has been a focus of maize research for a number of years \citep{Doebley2004}. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in  branching \citep{DoebleyStecGustus1995} and tillering \citep{DoebleyStec1991} differences between maize and teosinte.  A 4.9 kb retrotransposon (\emph{Hopscotch}) insertion into the upstream control region of \emph{tb1} in maize acts to enhance expression of \emph{tb1}, thus repressing lateral organ growth \citep{Doebley1997, Studer2011}.  Dating of the \emph{Hopscotch} retrotransposon suggests that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \citep{Studer2011}. The effects of the \emph{Hopscotch} insertion have been studied in maize \citep{Studer2011}, and analysis of teosinte alleles at \emph{tb1} has identified functionally distinct allelic classes \citet{StuderDoebley2012}, but little is known about the role of \emph{tb1} or the \emph{Hopscotch} insertion in natural populations of teosinte.

>>>>

The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of lateral organ growth, is a major QTL involved in branching differences between maize and its wild progenitor, teosinte. Previous work has shown that the insertion of a transposable element (\emph{Hopscotch}) upstream of \emph{tb1} enhances the gene's expression, causing much of the reduction in tillering observed in domesticated maize. Observations of the maize \emph{tb1} allele in teosinte individuals, coupled with estimates of an age of insertion of the \emph{Hopscotch} element that predates domestication, led us to investigate its prevalence and potential role in teosinte. Results from genotyping across many natural populations suggest that the \emph{Hopscotch} element is segregating at a higher than expected frequency in a number of populations of two subspecies of teosinte, \emph{Zea mays} ssp. \emph{parviglumis} and \emph{Zea mays} ssp. \emph{mexicana}. Analysis of linkage disequilibrium between the \emph{Hopscotch} element and variation in surrounding regions does not support a hypothesis of recent introgression from maize into teosinte.  Population genetic signatures, however, are consistent with selection on this locus and suggest the \emph{Hopscotch} insertion at \emph{tb1} may play an ecological role in teosinte. Finally, two greenhouse experiments with \emph{Zea mays} ssp. \emph{parviglumis} do not suggest \emph{tb1} controls tillering in natural populations of this subspecies. Our findings suggest that the role of the \emph{Hopscotch} in tillering in teosinte differs from domesticated maize, and that other loci may play a role in observed variation in this trait.

@@

Many morphological changes are associated with maize domestication, and understanding the genetic basis of these changes has been a focus of maize research for a number of years \citep{Doebley2004}. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in  branching \citep{DoebleyStecGustus1995} and tillering \citep{DoebleyStec1991} differences between maize and teosinte.  A 4.9 kb retrotransposon (\emph{Hopscotch}) insertion into the upstream control region of \emph{tb1} in maize acts to enhance expression of \emph{tb1}, thus repressing lateral organ growth \citep{Doebley1997, Studer2011}.  Dating of the \emph{Hopscotch} retrotransposon suggests that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \citep{Studer2011}. The effects of the \emph{Hopscotch} insertion have been studied in maize \citep{Studer2011}, and analysis of teosinte alleles at \emph{tb1} has identified functionally distinct allelic classes \citep{StuderDoebley2012}, but little is known about the role of \emph{tb1} or the \emph{Hopscotch} insertion in natural populations of teosinte.
